\chapter[Model updates]{Updates to the EMEP MSC-W model, 2020--2021}
\label{ch:ModelUpdates}

%% authors? added all names seen in Log.changes

{\bf{David Simpson, Michael Gauss, Qing Mu, Svetlana Tsyro and Peter Wind}}
\vspace{30pt}

%\old{TODO}
\COMMENT{STARTED Aug 2021. Needs to be filled in by authors}

This chapter summarises the changes made to the EMEP MSC-W  model
since \citet{R2020:ModDev}, and along with changes discussed in
\citet{R2013:ModDev,R2015:ModDev,R2016:ModDev,R2017:ModDev,R2019:ModDev,R2020:ModDev} and
\citet{R2014:ModDev},
updates the standard description given in \citet{Simpson_et_al:EMEP}. The
model version used for reporting this year is denoted rv4.42, which has
had some major updates (especially with regard to emissions) compared to
the rv4.35 reported in \citet{R2020:ModDev}.
Table~\ref{tab:Updates} summarises
the changes made in the EMEP model since the version documented in
\citet{Simpson_et_al:EMEP}, and these changes are discussed in
more detail in Sects.~\ref{sec:updateOverview}-\ref{sec:updateOther}.

%Versions:
%
%  rv4.35 used  for R2020
%  rv4.33 used  for R2019
%  rv4.33 - open source June 2019
%  rv4.32 used for EMEP course, April 2019
%  rv4.17a used for R2018 runs  (July-ish?)
%  rv4.17 released 26/2/2018
%  rv4.16 interim 21/12/2017 - used for N2O5 paper, wheat calculations
%  rv4.15 released 8/9/2017

\section{Overview of changes} 
\label{sec:updateOverview}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\begin{footnotesize}
\caption{Summary of major EMEP MSC-W model versions from 2012--2020.
Extends Table S1 of \citealt{Simpson_et_al:EMEP}
  COMMENT: Note - recent changes and tags are a little confused. Dave will fix.
}
\label{tab:Updates}
\centering
\begin{tabular}{lp{11cm}l}
\hline
Version & Update                                        & Ref$^{(a)}$   \\
\hline
rv4.42  & Changed default Kz scheme &  This report        \\
        & Updated emissions from CAMS81 project: DMS, soil NO;  &  This report        \\
% 21st May 2021:
rv4.41  & Updated emissions from CAMS81 project: DMS, soil NO; % 16/5
          Modified various parameters concerning sea-salt and wet-radius calculations; % 16/5
   fracPM25 modified;
          Revised global monthly factor;  % 16/5
        & \\
% 13/5
rv4.40  & Shipping emissions now spread as 20\% below 20m, 80\% between 20-90m;  
          New CAMS81-based aircraft emissions; % 29-30/4
          New CAMS81-based DMS emissions; % 29/4
        & \\
%SKIP BUG fix on aircraft emissions  27/4
% 21/4
rv4.39 &
  GNFR-CAMS is now the default emissions system - see Sect.~\ref{ssec:gnfr};  % 21/4
          Uses global time-zone map;   % 19/3
          Revised global monthly factor?? or rv4.41?;  % 19/3
          New default Kz method introduced (Sect.~\ref{ssec:PBL});
          NWP model rh2m now used in place of earlier sub-grid calculation; % 9/2
          Upgrade of local fractions methodology;  % 7/11/2020
        & \\
% 23/10/2021:
rv4.35  & Various updates, including heavy   
          refactoring of local-fraction code, bug-fixes in MARS module,
          and updates in chemical mechanisms, default PM and NMVOC speciation and
          GenChem systems     & R2020            \\
rv4.34  & Public domain (Feb. 2020); EmChem19a, EmChem19p      & R2020            \\
rv4.33  & Public domain (June 2019);
         EmChem19, PAR bug-fix, EQSAM4clim    & R2019            \\
rv4.32  & Used for EMEP course, April 2019    &    \\
rv4.30  & Moved to new GenChem-based system  &   \\
%        &                                    &        \\
rv4.17a & Used for R2018. Small updates         & R2018      \\
rv4.17  & Public domain (Feb. 2018);
         Corrections in global land-cover/deserts; added
          'LOTOS' option for European \ce{NH3} emissions; corrections
          to snow cover & R2018 \\
rv4.16  & New radiation scheme (Weiss\&Norman); Added dry and wet deposition for \ce{N2O5};
         (Used for  \citealt{Stadtler2018,MillsGCB2018b}) & R2018   \\
rv4.15  & EmChem16 scheme & R2017 \\
%    Sect.\ref{sec:GNFR}--\ref{ss:Splits} & R2016  \\
%
rv4.14  & Updated chemical scheme & R2017       \\
%% rv4.13 + CRI was used in McFiggans. Difficult to describe combo
%        & \\
rv4.12  & New  global land-cover and BVOC & R2017       \\
%        & \\
rv4.10  &  Public domain (Oct. 2016)                 
         (Used for  \citealt{MillsGCB2018a}) &  R2016 \\
%        & \\
rv4.9   & Updates for GNFR sectors, DMS, sea-salt, dust, \ce{S_A} and  $\gamma$, \ce{N2O5} & \\ 
rv4.8   &  Public domain (Oct. 2015); ShipNOx introduced.                          
         Used for EMEP HTAP2 model calculations, see
         special issue:
         \url{www.atmos-chem-phys.net/special_issue390.html},
          and \citet{Jonson_et_al:2017}.              & R2015\\
rv4.7   & Used for reporting, summer 2015;
         New calculations of aerosol surface area; 
         New gas-aerosol uptake and \ce{N2O5} hydrolysis rates; 
         Added 3-D calculations pf aerosol extinction and AODs;
         Emissions - new flexible mechanisms for interpolation and merging sources;
         Global - monthly emissions from ECLIPSE project;
         Global -  LAI changes from LPJ-GUESS model;
         WRF meteorology \citep{SkamarockKlemp2008} can now
     be used directly in EMEP model. & R2015 \\
%        &                                                &\\
rv4.6   & Used for Euro-Delta SOA runs                   & R2015  \\
%QUERY        & Bug-fix for ammonium deposition & \\
       & Revised boundary condition treatments % & \\  % Vertical profiles
       ; ISORROPIA capability added & \\
%       &                                                &\\
rv4.5  & Sixth open-source (Sep 2014);                    
        Improved dust, sea-salt, SOA modelling          % &      \\
       ; AOD and extinction coefficient calculations  updated %& \\
       ; Data assimilation system added % & \\
       ; Hybrid vertical coordinates replace earlier sigma % & \\
       ; Flexibility of grid projection increased. & R2014\\
%SKIP        & ?? Point sources, plume rise, data-assimilation\\
%       &                                                &\\
rv4.4   & Fifth open-source (Sep 2013) %
       ; Improved dust and sea-salt modelling   %          &      \\
       ; AOD and extinction coefficient calculations added %  &\\
       ; gfortran compatibility improved            %      &      \\
                  & R2014, R2013\\
%       &                                                &\\
rv4.3   & Fourth public domain (Mar. 2013)  %
       ; Initial use of namelists           %            & \\
       ; Smoothing of MARS results         %            & \\
       ; Emergency module for volcanic ash and other events% & \\
       ; Dust and road-dust options added as defaults % & \\
       ; Advection algorithm changed  % & \\ % \citet{CLAPP98}    & \\
             & R2013\\ 
%        &                                                &\\
rv4.0   & Third public domain (Sep. 2012), as \citet{Simpson_et_al:EMEP}            & R2013\\ 
%        & As documented in \citet{Simpson_et_al:EMEP}    & \\
%v2011-06& Second public domain (Aug. 2011)                &\\ 
%rv3     & First public domain (Sep. 2008)                &\\ 
        &                                                &\\
\hline
\end{tabular}
Notes: (a) R2018 refers to EMEP Status report 1/2018, etc.
\end{footnotesize}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{itemize}

\item Introduced new soil-NO, DMS and aircraft emissions, from CAMS-81 project
(Sect.~\ref{sec:updateEmis}).  -- Dave, Michael

\item Revised methods for vertical diffusion (Kz) \QUERY{and Hmix?} (Sect.~\ref{sec:updateKz}). -- Qing

\item Modified fine/coarse split of sea-salt (Sect.~\ref{ssec:updateSS}).

\item
Updated Local Fractions capabilities (Sect.~\ref{sec:updateLF}). -- Peter

\item  fracPM25 stuff (Sect.~\ref{sec:updateOther}). -- Dave
\item revised wet-radius calculations (Sect.~\ref{sec:updateOther}). --Dave

%\item The EMEP model's chemical pre-processing system and associated box-model (GenChem, boxChem) have been released as open-source.  See Sect.~\ref{sec:GChem}.

\item
Emissions speciation. New default and country-specific emission
speciations for NMVOC and \pmfine have been implemented.  See
Sect.~\ref{ssec:emissplits}.

\item
Numerous small changes to make the code more flexible and/or to
fix minor bugs.

\end{itemize}

In addition to these changes, articles on GenChem (EMEP's chemical pre-processing system)  and
the local-scale extension \COMMENT{Phrasing?} have now been published \citep{Simpson:GenChem,Denby:2020}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{New emission inputs}
\label{sec:updateEmis}

From CAMS81...

\subsection{New basis for emission sectors: GNFR\_CAMS}
\label{ssec:gnfr}

\COMMENT{Text/Table from gmdPaper}
 
Gridded anthropogenic emissions from CEIP
were previously categorized into 11
SNAP sectors, but for many years now EMEP emission reporting has been conducted and prepared for
modelling using the 13-sector
GNFR system\footnote{
GNFR=Gridding nomenclature for reporting/UNECE nomenclature for reporting of emissions to air,
e.g. \citealt{CEIP2020:GNFR}}.
In 2020 a 19-sector emission system (`GNFR\_CAMS') was implemented in the EMEP model, to take care of
emissions provided by TNO as part of the Copernicus CAMS project \citep{Kuenen2021}. This extended emissions system
enables for example four road traffic sectors, F1--F4, with e.g. F1
 representing exhaust emissions from gasoline vehicles.
Such emission sectors are characterised in the model by
release heights, timefactors and species-splits (e.g. NOx to NO and
\ce{NO2}, or NMVOC to individual VOC surrogates) for each sector.

Table~\ref{tab:GNFRsectors} summarises the new GNFR\_CAMS sectors,
and the mapping indices used.
%\COMMENT{Add index explanation in SI table?}


\begin{table}
\caption{The `GNFR\_CAMS' 19-sector system, and mapping indices for release heights, timefactors and species-splits,
 which is now
 default in EMEP model. \label{tab:GNFRsectors}
}
\begin{tabular}{lrrrrlc}
\hline
code & Number  & \multicolumn{3}{c}{Index}      & sector & SNAP \\ \cline{3-5}
     &         & timefac  & height & emissplit  &        & equivalent \\
\hline
A &      01 &   1 &  1 &   1 &  Public Power  &  1 \\
B &      02 &   3 &  3 &   2 &  Industry  & 3, 4 \\
C &      03 &   2 &  2 &   3 &  Other Stationary Combustion & 2 \\
D &      04 &   4 &  4 &   4 &  Fugitive & ??  \\
E &      05 &   6 &  2 &   5 &  Solvents & 6 \\
F &      06 &   7 &  2 &   6 &  Road Transport & 7 \\
G &      07 &   8 &  2 &   7 &  Shipping & 8  \\
H &      08 &   8 &  7 &   8 &  Aviation & 8  \\
I &      09 &   8 &  2 &   9 &  Offroad & 8  \\
J &     10 &   9 &  6 &  10 &  Waste & ? \\
K &     11 &  10 &  2 &  11 &  Agri - Livestock & 10 \\
L &     12 &  10 &  2 &  12 &  Agri - Other & 10 \\
M &     13 &   5 &  5 &  13 &  Other & 11 \\
A1 &    14 &   1 &  1 &   1 &  PublicPower - Point & 1 \\
A2 &    15 &   1 &  3 &   1 &  PublicPower - Area & 1 \\
F1 &    16 &   7 &  2 &  16 &  Road Transport - Exhaust Gasoline & 7  \\
F2 &    17 &   7 &  2 &  17 &  Road Transport - Exhaust Diesel& 7   \\
F3 &    18 &   7 &  2 &  18 &  Road Transport - Exhaust LPGgas & 7  \\
F4 &    19 &   7 &  2 &  19 &  Road Transport - NonExhaust Other& 7   \\
\hline
\end{tabular}
\end{table}



\subsection{Soil NO emissions}
\label{ssec:soilNO}

\COMMENT{DAVE to COMPLETE}

The EMEP model now makes uses of a new global dataset for soil NO emissions. This dataset, described in detail in \citep{SimpsonDarras2021}, 
provides gridded monthly data and the corresponding 3-hourly weight factors
at \ResEq{0.5} degrees horizontal resolution, over the period 2000-2018.  Emissions are provided as total values and also with separate data for soil
NO emissions from background biome values, and  those induced by fertilizers/manure, pulsing effects, and atmospheric
deposition, so that users can include, exclude or modify each component if wanted.

The basic methodology
merges methods from \citet{YiengerLevy:1995}
(hereafter YL95, c.f. Tab.~\ref{tab:Abbr}), with various updates to reflect recent literature
\citep[especially][hereafter SL11]{SteinkampLawrence2011}, and some simplifications
which reflect lack of  availability of some key data.
In YL95 and SL11, background biome emission factors ($A_{biome}$)  were modified by estimates of locally available nitrogen ($N_{avail}$),
which consists mainly of agricultural inputs of N (N from fertilizer, manure, hereafter \Nfert), or atmospheric deposition
of reactive nitrogen (hereafter \Ndep), and a pulse factor, \Fp. For this work we prefer to calculate the contributions of
\Nfert, \Ndep and \Fp separately, so we have:

\begin{equation}
  \Fs = \Fb  + \Ff + \Fd + \Fp
  \label{eqn:Fs}
\end{equation}

where the flux terms have units \ngN. The calculations of \Fb, \Ff, \Fd and \Fp are detailed in \citep{SimpsonDarras2021}.

An important issue arose during the construction ....
Within the LRTAP Convention,
most countries mainly report NOx emissions due to agricultural activities
using the EMEP/EEA Emissions Inventory Guidebook
\citep{Guidebook2019:3D}. The Guidebook provides methods for calculating
soil-NO data from fertilizer and other inputs.

Table~\ref{tab:GB2019} presents the main sources for which
 soil NO emissions ares covered by the Guidebook, and
Table~\ref{tab:Res2015} presents the nationally
submitted emissions following this system (data
provided by Sabine Schindlbacher, EMEP CEIP, 2021).
It can be seen that for the vast majority of countries
the main emission categories are 3Da1,3Da2a-c,and
3Da3. These are all roughly equivalent to
the `Fert' emissions from CAMS-GLOB-SOIL.

When using EMEP emissions derived from
officially reported data (with soil NO emissions as given in GNFR L),
for example in EMEP MSC-W reporting runs, retain the
official GNFR L data, but
 but add biome, N-dep and pulse emissions from CAMS-GLOB-SOIL.


\citet{SimpsonDarras:2021} discuss these double-counting issues for
ECLIPSE v5 and v6 inventories, and for CAMS-GLOB and CAMS-REG. In
some cases the CAMS-GLOB-SOIL \Nfert emissions should be used, but
in other cases not. In the EMEP model this ....



 TO BE CONTINUED!!!!



\subsection{DMS emissions}
\label{ssec:DMS}

\COMMENT{MICHAEL to COMPLETE}

A new method to calculate biogenic emissions of dimethyl sulphide (DMS) has been added to the EMEP model. It is based on the climatology of sea surface DMS concentrations published by \citet{Lana2011} and the meteorological data driving the EMEP model, and it has been used for this year's status run. 

Using the sea surface DMS concentrations from \citet{Lana2011}, the fluxes of DMS from the water into the atmosphere are calculated online based on .

 and
calculates the Schmidt number ($Sc$) from the sea surface temperature ($SST$, given in \textdegree C) as
\\
calculates $Sc$ with the formula of \citet{Wanninkhof2014} and $K_{w}$ :
\\
\\
$Sc=2855.7 -177.63\  SST + 6.0438\  SST^2 - 0.11645\  SST^3 + 0.00094743\ SST^4$ 

The transfer velocity $K_{w}$ (given in cm hour$^{-1}$) is calculated with the formula of \citet{Nightingale2000}, depending on the 10-meter wind speed $u_{10}$ (given in \ms):
\\
$K_w=(0.222\  u_{10}^2 +0.333\ u_{10}) \sqrt{\frac{600}{Sc}} $\\


Following \citet{Tarrason:1995} we assume that 66\% of the DMS in air is converted into \ce{SO2}; the DMS emission is thus implemented in the model as an \ce{SO2} flux as:

\begin{verbatim}
flux(SO2)=0.66*O_DMS*Kw/3600*0.01*GRAV*roa/p*AVOG  
\end{verbatim}

where $O\_DMS$ is the sea surface DMS concentration (in nanomol L$^{-1}$),
$GRAV$ is standard gravity (9.807 m s$^{-2}$),
$roa$ is the ambient density of air (kg m$^{-3}$),
$p$ is ambient air pressure
and $AVOG$ is Avogradro's number (6.023e+23 mol$^{-1}$).

\subsection{Aircraft emissions}
\label{ssec:Aircraft}

\COMMENT{MICHAEL to COMPLETE}

LTO (landing/take-off) emission data are, as before, taken from CEIP, i.e. based on aircraft emissions officially reported by countries for up to 3000 feet above the surface. However, aircraft emissions at higher altitudes are now based on the new CAMS-GLOB-AIR dataset (\citet{GranierCAMS2019}), given separately for all years from 2000 to 2019, with monthly variation and a spatial resolution of 0.5$^\circ$\,$\times$\,0.5$^\circ$\,$\times$\,610\,m. These data have been provided through the Copernicus Atmosphere Monitoring Service and can be downloaded from the Atmosphere Data Store of ECMWF (\url{https://ads.atmosphere.copernicus.eu/}).

We assume that, in approximation, half of the LTO emissions (from CEIP) occur below 50 meters above the surface, while the other half is evenly distributed over the height range between 50 meters and about 3000 feet.

In the EMEP model, CAMS-GLOB-AIR emissions between 0 and 610m (2000 ft) above the surface are not used, and half of the CAMS-GLOB-AIR emissions between 610 and 1220m are evenly distributed over the model layers corresponding to that height range. The assumption is that emissions up to 3000 feet should already be fully included in the CEIP (LTO) emissions. Above 1220m (4000ft), CAMS-GLOB-AIR emissions are used in full, and interpolated to the relevant model grid during model runtime.

The option of using the old QUANTIFY data (website \url{www.pa.op.dlr.de/quantify}) instead of CAMS-GLOB-AIR is still available.

\subsection{Revised fine/coarse splits of sea-salt emissions}
\label{ssec:updateSS}

The split of sea salt between the fine and coarse fractions have been revised based on the results of comparison of modelled \chem{Na+} in \PM[2.5] and \PM[10] with EMEP observations. Namely, the fraction of sea salt emitted in \PM[2.5] fraction has been slightly increased (while the total sea salt emissions in \PM[10] remain unchanged). 

More specifically, sea salt is emitted in seven size bins up to 10 \um in diameter, from which the smallest four bins were previously assigned to the \PM[2.5] fraction and the remaining three bins comprised coarse sea salt. Now, the fifth bin is split 50/50 \% between the \PM[2.5] and \PM[2.5-10] fractions.  


This modification has a minor effect on formation of coarse \noiii as the surface area of coarse sea salt has slightly decreased.
\COMMENT{SVETLANA to COMPLETE}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Emission speciation}
\label{ssec:emissplits}.

\COMMENT{OLD:}

The emissions speciation of NMVOC and primary PM (PPM) were updated
for the EmChem19a scheme to reflect recent data available from
the latest TNO/CAMS inventories (see chap.~\ref{chap:TNO} 
\COMMENT{Dave will sort out cross-refs}
, also
\citealt{CAMSemis2019}).  For NMVOC the main changes have been:

\begin{enumerate}
  \item Use TNO NMVOC speciation

    TNO provided NMVOC speciation data for 25 compounds from each
    GNFR\_CAMS category, as part of the CAMS-REG-v3.1.2 database,
    which were then mapped to the EmChem19a species.

  \item Improve country-specific road transport estimates

    TNO provided country-specific fractions of the four road-traffic
    sectors in the CAMS-REG-AP\_v2.2.1 inventory for 2015, specifically
    F1=gasoline exhaust, F2=diesel exchaust, F3=LPG exhaust, and
    F4=non-exhaust emissions. These were also aggregated to form
    country-specific NMVOC splits for the generic GNFR F category
    (road transport).

\end{enumerate}

% PM TNO CAMS-REG_AP_v2.2.1_2015_REF

For \pmfine and \pmten, we made use of country-data generated by TNO for
the TFMM Euro-DeltaCarb simulations, which gave EC, OM (named
OC in the files), Na, SO4 and other compounds, as well as the fraction
of modern carbon in the GNFR-C categories. These data were aggregated
to the EMEP models EC, OM, and remPPM compounds. Data were provided for
both the Ref1 and Ref2 cases (see Chap.~\ref{chap:TNO}), and we assume
that Ref1 data are appropriate for modelling with officially submitted
emissions, and Ref2 data appropriate for modelling when Ref2 emissions
are used (e.g. for the \textbf{EMEPwRef2C} simulations discussed in
Chapters~\ref{ch:chapterStatus}--\ref{ch:Condensables}).
\COMMENT{Dave will sort out text later}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Revised PBL parametrisations}
\label{sec:updateKz}

\COMMENT{QING to UPDATE}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Local Fractions}
\label{sec:updateLF}
The range of applications of the Local Fractions has been extended. 

The sources can be defined by either a list of countries or group of countries, as small squares surrounding the receptor grid (with user defined square sizes), or as regions defined as "masks" given in a NetCDF file.

Wet and dry deposition (separately) per source country can now be computed.

\COMMENT{PETER to UPDATE}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Other?}
\label{sec:updateOther}

fracMPM25 - Dave
revised wet-radius calculations --Dave

Something on pollen perhaps? -- Alvaro??



\clearpage
\bibliographystyle{copernicus}         % change bibliography-name after each
\renewcommand\bibname{References}      % bibliographystyle command!
\addcontentsline{toc}{section}{References}
\bibliography{Refs,EMEP_Reports,uEMEP,RefsUpdates}
